---
title: "2.D: PANTHER & rbioapi"
author: "Moosa Rezwani"
description: >
 Connect to PANTHER in R with rbioapi package. 
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    anchor_sections: true
    number_sections: true
    self_contained: true
    dev: png
    encoding: 'UTF-8'
    
vignette: >
  %\VignetteIndexEntry{2.D: PANTHER & rbioapi}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r message=FALSE, include=FALSE, results="hide", setup, echo=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  tidy = FALSE,
  cache = FALSE,
  dev = "png",
  comment = "#>"
)
library(rbioapi)
rba_options(timeout = 30, skip_error = TRUE)
```

# Introduction {#introduction}

To describe PANTHER in its authors’ own words, we quote directly from Thomas et al. (2022):

> Phylogenetics is a powerful tool for analyzing protein sequences, by inferring their evolutionary relationships to other proteins. However, phylogenetics analyses can be challenging: they are computationally expensive and must be performed carefully in order to avoid systematic errors and artifacts. Protein Analysis THrough Evolutionary Relationships (PANTHER; http://pantherdb.org) is a publicly available, user-focused knowledgebase that stores the results of an extensive phylogenetic reconstruction pipeline that includes computational and manual processes and quality control steps. [...] The PANTHER knowledgebase can be downloaded or accessed via an extensive API. In addition, PANTHER provides software tools to facilitate the application of the knowledgebase to common protein sequence analysis tasks: exploring an annotated genome by gene function; performing "enrichment analysis" of lists of genes; annotating a single sequence or large batch of sequences by homology; and assessing the likelihood that a genetic variant at a particular site in a protein will have deleterious effects.
>
> (source: Thomas PD, Ebert D, Muruganujan A, Mushayahama T, Albou L-P, Mi H. "PANTHER: Making genome-scale phylogenetics accessible to all." *Protein Science* 31.1 (2022): 8--22. <https://doi.org/10.1002/pro.4218>)

The available tools in PANTHER's **RESTful API services** can be divided into 3 broad categories: Mapping genes, retrieving information, and research tools. Herein, we provide a very short introduction; you can always check functions' manuals for detailed guides and examples.

------------------------------------------------------------------------

# Map genes {#map-genes}

-   `rba_panther_mapping()`: map your gene-set to PANTHER database and retrieve attributes and annotations associated with your genes

-   `rba_panther_ortholog()`: Retrieve Orthologs of your genes

-   `rba_panther_homolog()`: Retrieve Homologs of your genes

------------------------------------------------------------------------

# Get information {#get-information}

-   `rba_panther_info()`: Retrieve a list of PANTHER's supported organisms, datasets, families, or pathways

-   `rba_panther_family()`: Retrieve Orthologs, MSA, or Tree topology of a given PANTHER family.

------------------------------------------------------------------------

# Gene List Analysis {#gene-list-analysis}

`rba_panther_enrich()` is equivalent to [Gene List analysis tool's webpage](https://www.pantherdb.org/index.jsp "PANTHER Gene List Analysis"). Depending on the provided input's class, PANTHER will perform either over-representation analysis or statistical enrichment analysis. Below we demonstrate how to perform such analyses.

## Get the available annotation datasets {#analysis-get-available-annotation-datasets}

First, we need to select an annotation dataset to conduct the analysis based on it. Each annotation dataset contains a collection of terms, where each term is associated with a group of genes.

To retrieve the list of available annotation datasets in PANTHER, use the following command:

```{r enrich_available_annotations}
annots <- rba_panther_info(what = "datasets")
```

```{r enrich_available_annotations_results, echo=FALSE}
if (is.data.frame(annots)) {
  DT::datatable(
    data = annots,
    options = list(
      scrollX = TRUE, 
      paging = TRUE,
      fixedHeader = TRUE,
      keys = TRUE
    )
  )
} else {
  print("Vignette building failed. It is probably because the web service was down during the building.")
}
```

Please note that you should use the ID of the desired annotation dataset, not its label. For example, using `"biological_process"` is incorrect; you should rather use `"GO:0008150"`.

## Submit the analysis request {#submit-the-analysis-request}

Depending on the provided input, PANTHER will conduct two types of analysis:

1.  If a character vector is supplied, over-representation analysis will be performed using either Fisher's exact or binomial test.

2.  If a data frame with gene identifiers and their corresponding expression values is supplied, statistical enrichment test is performed using Mann-Whitney U (Wilcoxon Rank-Sum) test.

rbioapi determines the proper analysis based on the class of the `genes` parameter. Please refer to the details section of `rba_panther_enrich()` function manual for more information.

### Over-representation analysis {#over-representation-analysis}

Now, suppose we want to perform an over-representation analysis against the 'GO biological process' annotation dataset. In this example, we only provide the gene names, thus over-representation analysis will be conducted:

```{r rba_panther_overrep, message=TRUE}
# Create a variable to store the genes vector
my_genes_vec <- c(
  "p53", "BRCA1", "cdk2", "Q99835", "CDC42", "CDK1","KIF23","PLK1",
  "RAC2","RACGAP1","RHOA", "RHOB", "PHF14", "RBM3", "MSL1"
)

# Submit the analysis request.
enriched <- rba_panther_enrich(
  genes = my_genes_vec,
  organism = 9606,
  annot_dataset = "GO:0008150",
  cutoff = 0.05
)

# Note that we did not supply the `test_type` parameter.
# In this case, the function defaults to Fisher's exact test
# (`test_type = "FISHER"`).
# You may also use the binomial test for over-representation analysis
# (`test_type = "BINOMIAL"`).
```

```{r rba_panther_overrep_results, echo=FALSE}
if (utils::hasName(enriched, "result") && is.data.frame(enriched$result)) {
  DT::datatable(
    data = enriched$result,
    options = list(
      scrollX = TRUE, 
      paging = TRUE,
      fixedHeader = TRUE,
      keys = TRUE,
      pageLength = 10
    )
  )
} else {
  print("Vignette building failed. It is probably because the web service was down during the building.")
}
```

### Statistical enrichment analysis {#statistical-enrichment-analysis}

As you can see in the above example, only a vector of gene names was used. We can also use the corresponding expression values of the genes. In this case, PANTHER will perform a statistical enrichment analysis.

To do so, the only change will be to supply a data frame to the `genes` parameter. Note that in this case, Mann-Whitney U Test will be performed. The data frame should have two columns: the first column should contain the gene identifiers as a character vector; the second column should contain the corresponding expression values as a numeric vector.

```{r rba_panther_enrich, eval=FALSE}
# Create a variable to store the data frame
my_genes_df <- data.frame(
  genes = c(
    "p53", "BRCA1", "cdk2", "Q99835", "CDC42", "CDK1","KIF23","PLK1",
    "RAC2","RACGAP1","RHOA", "RHOB", "PHF14", "RBM3", "MSL1"
  ),
  ## generate random expression values
  expression = runif(15, 0, 10) 
)

# Submit the analysis request.
enriched <- rba_panther_enrich(
  genes = my_genes_df,
  organism = 9606,
  annot_dataset = "GO:0008150",
  cutoff = 0.05
)

# Note that we did not supply the `test_type` parameter.
# In this case, the function defaults to the Mann-Whitney U test
# (`test_type = "Mann-Whitney"`).
# This is the only valid value for statistical enrichment analysis,
# so omitting or supplying it makes no difference.
```

**Please Note:** Other services supported by rbioapi also provide Over-representation analysis tools. Please see the vignette article [Do with rbioapi: Over-Representation (Enrichment) Analysis in R](rbioapi_do_enrich.html) ([link to the documentation site](https://rbioapi.moosa-r.com/articles/rbioapi_do_enrich.html)) for an in-depth review.

------------------------------------------------------------------------

# Tree grafter {#tree-grafter}

`rba_panther_tree_grafter()` is an equivalent to the "[Graft sequence into PANTHER library of trees](https://www.pantherdb.org/tools/sequenceSearchForm.jsp)" tool.

------------------------------------------------------------------------

# How to Cite? {#citations}

To cite PANTHER (Please see <https://www.pantherdb.org/publications.jsp#HowToCitePANTHER>):

-   Thomas PD, Ebert D, Muruganujan A, Mushayahama T, Albou L-P, Mi H. PANTHER: Making genome-scale phylogenetics accessible to all. *Protein Science*. 2022;31(1):8--22. <https://doi.org/10.1002/pro.4218>

For enrichment and over-representation analyses, also cite:

-   Mi H, Muruganujan A, Huang X, Ebert D, Mills C, Guo X, Thomas PD. Protocol Update for large-scale genome and gene function analysis with the PANTHER classification system (v.14.0). *Nature Protocols*. 2019;14:703--721. <https://doi.org/10.1038/s41596-019-0128-8>

For TreeGrafter, also cite:

-   Tang H, Finn RD, Thomas PD. TreeGrafter: phylogenetic tree-based annotation of proteins with Gene Ontology terms and other annotations. *Bioinformatics*. 2019;35(3):518--520. <https://doi.org/10.1093/bioinformatics/bty625>

To cite rbioapi:

-   Moosa Rezwani, Ali Akbar Pourfathollah, Farshid Noorbakhsh, rbioapi: user-friendly R interface to biologic web services' API, Bioinformatics, Volume 38, Issue 10, 15 May 2022, Pages 2952--2953, <https://doi.org/10.1093/bioinformatics/btac172>

------------------------------------------------------------------------

# Links {#links}

-   [This article in rbioapi documentation site](https://rbioapi.moosa-r.com/articles/rbioapi_panther.html "2.D: PANTHER & rbioapi")

-   [Functions references in rbioapi documentation site](https://rbioapi.moosa-r.com/reference/index.html#section-panther-rba-panther- "rbioapi reference")

-   [rbioapi vignette index](rbioapi.html "rbioapi: User-Friendly R Interface to Biologic Web Services' API")

------------------------------------------------------------------------

# Session info {#session-info}

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
