---
title: "2.G: UniProt & rbioapi"
author: "Moosa Rezwani"
description: >
 Connect to UniProt in R with rbioapi package.
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    anchor_sections: true
    number_sections: true
    self_contained: true
    dev: png
    encoding: 'UTF-8'

vignette: >
  %\VignetteIndexEntry{2.G: UniProt & rbioapi}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r message=FALSE, include=FALSE, results="hide", setup, echo=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  tidy = FALSE,
  cache = FALSE,
  dev = "png",
  comment = "#>"
)
library(rbioapi)
rba_options(timeout = 30, skip_error = TRUE)
```


# Introduction {#introduction}

Directly quoting from [UniProt](https://www.uniprot.org "Universal Protein Resource (UniProt)"):

> The Universal Protein Resource (UniProt) is a comprehensive resource for protein sequence and annotation data. The UniProt databases are the [UniProt Knowledgebase (UniProtKB)](https://www.uniprot.org/help/uniprotkb), the [UniProt Reference Clusters (UniRef)](https://www.uniprot.org/help/uniref), and the [UniProt Archive (UniParc)](https://www.uniprot.org/help/uniparc). The UniProt consortium and host institutions EMBL-EBI, SIB and PIR are committed to the long-term preservation of the UniProt databases.
>
> (source: <https://www.uniprot.org/help/about>)

------------------------------------------------------------------------

# Search vs Retrieve {#search-vs-retrieve}

Most rbioapi UniProt function families have two variants: one retrieves data using a supported identifier, while the other (with a `_search` suffix) searches using one or more service-specific criteria. We first demonstrate this using an example, then provide a list of these function families in rbioapi.

Suppose we are interested in the human CD40 ligand protein and know that its UniProt accession is "P29965". We can retrieve it directly:

```{r rba_uniprot_proteins}
## 1 Retrieve CD40 ligand by its UniProt accession
cd40 <- rba_uniprot_proteins(accession = "P29965")

## 2 Inspect the returned object's structure
str(cd40, 1)
```

This returns the data represented by the [UniProtKB entry for P29965](https://www.uniprot.org/uniprotkb/P29965/entry "UniProtKB - P29965 (CD40L_HUMAN)"). If we do not know the accession, or want to query by other criteria, we can use the function with the `_search` suffix:

```{r rba_uniprot_proteins_search}
## 1 Supply the criteria pertinent to the search
cd40_search <- rba_uniprot_proteins_search(
  protein = "CD40 ligand",
  organism = "human",
  reviewed = TRUE
)

## 2 Inspect the returned object's structure
str(cd40_search, 2)
```

This is analogous to an [advanced search](https://www.uniprot.org/help/advanced_search "UniProtKB advanced search options") in the UniProt web portal. See the `rba_uniprot_proteins_search` manual for details. In `*_search` functions you do not need to fill every argument, but the Proteins API requires at least one of the primary search criteria documented for that function; refining modifiers do not constitute a search by themselves.

The applications of `*_search` variants are not limited to what the title 'search' implies. These functions also retrieve their search hits, so you can use them for bulk retrieval. Many arguments accept multiple values; see the functions' manuals for the supported limits. Consider the following examples:

```{r rba_uniprot_proteins_search1_2}
## 1 As the simplest scenario, we can retrieve multiple proteins in one call
multi_prs1 <- rba_uniprot_proteins_search(
  accession = c("P04637", "P38398", "P24941", "P60953", "P06493", "Q02241")
)
## Inspect the returned object's structure
str(multi_prs1, 1)

## 2 Search by gene name, excluding isoforms and unreviewed entries
multi_prs2 <- rba_uniprot_proteins_search(
  gene = c("KIF23", "BRCA1", "TP53", "CDC42"),
  reviewed = TRUE,
  taxid = 9606,
  isoform = 0
)

str(multi_prs2, 1)
```

```{r rba_uniprot_proteins_search3}
## 3 Search for every protein with the chemokines keyword
multi_prs3 <- rba_uniprot_proteins_search(
  keyword = "chemokines"
)

str(multi_prs3, 1)
```

```{r rba_uniprot_proteins_search4}
## 4 Search for every reviewed SARS-CoV-2 protein
multi_prs4 <- rba_uniprot_proteins_search(
  organism = "SARS-CoV-2",
  reviewed = TRUE
)

str(multi_prs4, 1)
```

------------------------------------------------------------------------

# Functions with `*_search` variant {#functions-with-star-search-variant}

The search variants are not limited to the examples above. The following function families have both retrieval and search variants. See their manuals for detailed guidance and examples.

1.  `rba_uniprot_proteins()` and `rba_uniprot_proteins_search()`

2.  `rba_uniprot_features()` and `rba_uniprot_features_search()`

3.  `rba_uniprot_variation()` and `rba_uniprot_variation_search()`

4.  `rba_uniprot_antigens()` and `rba_uniprot_antigens_search()`

5.  `rba_uniprot_epitope()` and `rba_uniprot_epitope_search()`

6.  `rba_uniprot_rna_edit()` and `rba_uniprot_rna_edit_search()`

7.  `rba_uniprot_proteomes()` and `rba_uniprot_proteomes_search()`

8.  `rba_uniprot_mutagenesis()` and `rba_uniprot_mutagenesis_search()`

9.  `rba_uniprot_proteomics_hpp()` and `rba_uniprot_proteomics_hpp_search()`
 
10.  `rba_uniprot_proteomics_non_ptm()` and `rba_uniprot_proteomics_non_ptm_search()`
 
11.  `rba_uniprot_proteomics_ptm()` and `rba_uniprot_proteomics_ptm_search()`

12.  `rba_uniprot_genecentric()` and `rba_uniprot_genecentric_search()`

13.  `rba_uniprot_uniparc()` and `rba_uniprot_uniparc_search()`

------------------------------------------------------------------------

# UniProt functions categories {#uniprot-functions-categories}

The Proteins API endpoints implemented by rbioapi are organized into six groups. Here are those categories and the corresponding rbioapi functions. See the functions' manuals for more details.

## Proteins {#proteins}

### Proteins: {#proteins-proteins}

-   `rba_uniprot_proteins()`

-   `rba_uniprot_proteins_search()`

-   `rba_uniprot_proteins_crossref()`

### Features {#proteins-features}

-   `rba_uniprot_features()`

-   `rba_uniprot_features_search()`

### Variation {#proteins-variation}

-   `rba_uniprot_variation()`

-   `rba_uniprot_variation_locations()`

-   `rba_uniprot_variation_search()`

### Antigens {#antigens}

-   `rba_uniprot_antigens()`

-   `rba_uniprot_antigens_search()`

### Epitopes {#epitopes}

-   `rba_uniprot_epitope()`

-   `rba_uniprot_epitope_search()`

### Mutagenesis {#mutagenesis}

-   `rba_uniprot_mutagenesis()`

-   `rba_uniprot_mutagenesis_search()`

### RNA-Editing {#rna-editing}

-   `rba_uniprot_rna_edit()`

-   `rba_uniprot_rna_edit_search()`

## Proteomics {#proteomics}

-   `rba_uniprot_proteomics_species()`

-   `rba_uniprot_proteomics_hpp()`
 
-   `rba_uniprot_proteomics_hpp_search()`
 
-   `rba_uniprot_proteomics_non_ptm()`
 
-   `rba_uniprot_proteomics_non_ptm_search()`
 
-   `rba_uniprot_proteomics_ptm()`
 
-   `rba_uniprot_proteomics_ptm_search()`

## Proteomes {#proteomes}

-   `rba_uniprot_proteomes()`

-   `rba_uniprot_proteomes_search()`

-   `rba_uniprot_genecentric()`

-   `rba_uniprot_genecentric_search()`

## Taxonomy {#taxonomy}

-   `rba_uniprot_taxonomy()`

-   `rba_uniprot_taxonomy_lca()`

-   `rba_uniprot_taxonomy_lineage()`

-   `rba_uniprot_taxonomy_name()`

-   `rba_uniprot_taxonomy_path()`

-   `rba_uniprot_taxonomy_relationship()`

## Coordinates {#coordinates}

-   `rba_uniprot_coordinates()`

-   `rba_uniprot_coordinates_search()`

-   `rba_uniprot_coordinates_location()`

-   `rba_uniprot_coordinates_location_protein()`

-   `rba_uniprot_coordinates_location_genome()`

## UniParc {#uniparc}

-   `rba_uniprot_uniparc()`

-   `rba_uniprot_uniparc_search()`

-   `rba_uniprot_uniparc_bestguess()`

-   `rba_uniprot_uniparc_sequence()`

------------------------------------------------------------------------

# How to Cite? {#citations}

To cite UniProt (Please see <https://www.uniprot.org/help/publications>):

-   The UniProt Consortium. (2025). UniProt: the Universal Protein Knowledgebase in 2025. *Nucleic Acids Research*, *53*(D1), D609–D617. <https://doi.org/10.1093/nar/gkae1010>
-   Nightingale, A., Antunes, R., Alpi, E., Bursteinas, B., Gonzales, L., Liu, W., Luo, J., Qi, G., Turner, E., & Martin, M. (2017). The Proteins API: Accessing key integrated protein and genome information. *Nucleic Acids Research*, *45*(W1), W539–W544. <https://doi.org/10.1093/nar/gkx237>

To cite rbioapi:

-   Moosa Rezwani, Ali Akbar Pourfathollah, Farshid Noorbakhsh, rbioapi: user-friendly R interface to biologic web services' API, Bioinformatics, Volume 38, Issue 10, 15 May 2022, Pages 2952--2953, <https://doi.org/10.1093/bioinformatics/btac172>

------------------------------------------------------------------------

# Links {#links}

-   [This article on the rbioapi documentation site](https://rbioapi.moosa-r.com/articles/rbioapi_uniprot.html "2.G: UniProt & rbioapi")

-   [Functions references in rbioapi documentation site](https://rbioapi.moosa-r.com/reference/index.html#section-uniprot-rba-uniprot- "rbioapi reference")

-   [rbioapi vignette index](rbioapi.html "rbioapi: User-Friendly R Interface to Biologic Web Services' API")

------------------------------------------------------------------------

# Session info {#session-info}

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
