---
title: "2.B: JASPAR & rbioapi"
author: "Moosa Rezwani"
description: >
 Connect to JASPAR in R with rbioapi package. 
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    anchor_sections: true
    number_sections: true
    self_contained: true
    dev: png
    encoding: 'UTF-8'

vignette: >
  %\VignetteIndexEntry{2.B: JASPAR & rbioapi}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r message=FALSE, include=FALSE, results="hide", setup, echo=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  tidy = FALSE,
  cache = FALSE,
  dev = "png",
  comment = "#>"
)
library(rbioapi)
rba_options(timeout = 30, skip_error = TRUE)
```


# Introduction {#introduction}

[JASPAR](https://jaspar.elixir.no) is an open-access database of curated transcription factor binding profiles. JASPAR 2026 is the 11th release. It expands the CORE and UNVALIDATED position frequency matrix collections and introduces a deep-learning collection containing BPNet models and their interpreted binding profiles.

> Baydar Ovek D, et al. JASPAR 2026: expansion of transcription factor binding profiles and integration of deep learning models. *Nucleic Acids Research*. 2026;54(D1):D184-D193; doi: [10.1093/nar/gkaf1209](https://doi.org/10.1093/nar/gkaf1209)

------------------------------------------------------------------------

# Data Organization in JASPAR {#data-organization-in-jaspar}

JASPAR is a database of transcription factor binding matrices with annotations and metadata. These entities are organized in a hierarchical fashion that we will explore next.

## Releases {#releases}

In addition to the latest JASPAR database release (2026), other active releases are also available. Most rbioapi JASPAR functions have a `release` argument; they use release 2026 by default.

```{r release}
## Call the function without any arguments to get a list of releases
releases <- rba_jaspar_releases()

## Supply a release number for details:
release_11_info <- rba_jaspar_releases(11)
```

## Collections {#collections}

Within a release, matrix profiles are organized into collections. You can use `rba_jaspar_collections()` to get a list of available collections, or read the "JASPAR Collections" section on the [JASPAR documentation website](https://jaspar.elixir.no/docs/ "JASPAR Documentation") for a thorough review.

```{r collections}
## To get a list of available collections in release 2026:
rba_jaspar_collections(release = 2026)


## You can list information on all matrices available in a collection:
mat_in_core_2026 <- rba_jaspar_collections_matrices(collection = "CORE")
```

## Taxonomic Groups {#taxonomic-groups}

Within each collection, the matrix profiles are organized based on main taxonomic groups:

```{r taxons}
## To get a list of taxonomic groups in release 2026:
rba_jaspar_taxons(release = 2026)


## You can list information on all matrices available in a taxonomic group:
mat_in_insects <- rba_jaspar_taxons_matrices(tax_group = "insects")
```

## Species {#species}

As we go down in the data organization hierarchy, each taxonomic group consists of species:

```{r species}
## To get a list of species in release 2026:
species <- rba_jaspar_species(release = 2026)
head(species)

## You can list information on all matrices available for a species:
mat_in_human <- rba_jaspar_species_matrices(tax_id = 9606)
```

------------------------------------------------------------------------

# Matrix Profiles {#matrix-profiles}

## Search Matrix Profiles {#search-matrix-profiles}

Retrieving a list of every matrix available in a given category is not the only option. You can also build a search query using `rba_jaspar_matrix_search`. Note that this is a search function, you are not required to fill every argument. You may use any combination of arguments you see fit to build your query. You can even call the function without any argument to get a list of all the matrix profiles. For instance:

```{r search, eval=FALSE}
## Get a list of all the available matrix profile:
all_matrices <- rba_jaspar_matrix_search()

## Search FOX:
FOX_matrices <- rba_jaspar_matrix_search(term = "FOX")

## Transcription factors named FOXP3
FOXP3_matrices <- rba_jaspar_matrix_search(term = "FOXP3")

## Transcription factors of Zipper-Type Class
zipper_matrices <- rba_jaspar_matrix_search(tf_class = "Zipper-Type")

## Transcription factors of Zipper-Type Class in PBM collection
zipper_pbm_matrices <- rba_jaspar_matrix_search(
  tf_class = "Zipper-Type",
  collection = "PBM"
)
```

## List Matrix Profiles Associated to a Base identifier {#list-matrix-profiles-associated-to-a-base-identifier}

Since JASPAR release 2010, matrix profiles have been versioned. A matrix profile identifier follows a "base_id.version" naming scheme; for example, "MA0600.2" corresponds to the second version of a matrix with base ID MA0600. You can use `rba_jaspar_matrix_versions` to get a list of matrix profiles with a given base ID. Also note that functions used to list available matrices generally have an argument called `only_last_version`.

```{r versions, eval=FALSE}
## Get matrix profiles versions associated to a base id
MA0600_versions <- rba_jaspar_matrix_versions("MA0600")
```

## Get a Matrix Profile {#get-a-matrix-profile}

Now that you listed or searched for matrix profiles, you can use `rba_jaspar_matrix` to retrieve matrix profiles. There are two ways in which you can use this function:

### Get Matrix and Annotations as an R Object {#get-matrix-and-annotations-as-an-r-object}

To do that, only fill in the `matrix_id` argument in `rba_jaspar_matrix`

```{r matrix_r_object}
pfm_matrix <- rba_jaspar_matrix(matrix_id = "MA0600.2")

## you can find the matrix in the pfm element along with
## other elements which correspond to annotations and details
str(pfm_matrix)
```

### Save a Matrix as a File in a Specific Format {#save-a-matrix-a-file-in-specific-format}

JASPAR provides position frequency matrices (PFM) formatted as **Raw PFM**, **JASPAR**, **TRANSFAC**, **YAML**, and **MEME**. You can download a matrix profile as a file with any of these formats. To do that, you should use the `file_format` and `save_to` arguments available in `rba_jaspar_matrix`. There are two notes here:

1.  In this case, the function saves your matrix as a file and returns the unparsed content as a character string.

2.  The `save_to` argument in this and other rbioapi functions can be used in several ways:\
    2.1. `save_to = NA`: rbioapi automatically generates a file path under your working directory, saves the file there, and reports the path in a message.\
    2.2. `save_to = file_name` without a directory: rbioapi saves the file with the supplied name in your working directory.\
    2.3. `save_to = directory_path`: rbioapi saves the file with an appropriate name in that directory.\
    2.4. `save_to = file_path`: rbioapi saves the file to the exact path. Ensure that the path's extension matches the requested file format; otherwise, rbioapi uses the supplied extension and issues a warning.

    In any of the aforementioned cases, the file path can be absolute or relative.

```{r matrix_save, eval=FALSE}
## Different ways in which you can save the matrix file:
meme_matrix1 <- rba_jaspar_matrix(
  matrix_id = "MA0600.2",
  file_format = "meme"
)

meme_matrix2 <- rba_jaspar_matrix(
  matrix_id = "MA0600.2",
  file_format = "meme",
  save_to = "my_matrix.meme"
)

meme_matrix3 <- rba_jaspar_matrix(
  matrix_id = "MA0600.2",
  file_format = "meme",
  save_to = "c:/rbioapi"
)

meme_matrix4 <- rba_jaspar_matrix(
  matrix_id = "MA0600.2",
  file_format = "meme",
  save_to = "c:/rbioapi/my_matrix.meme"
)
```

## Get Binding Sites for a Matrix Profile {#get-binding-sites-of-a-matrix-profiles}

If available, you can retrieve information on binding sites associated with a matrix profile. The result includes a data frame of genomic coordinates, URLs to FASTA and BED files, and other annotations.

```{r sites, eval=FALSE}
## Get binding sites for a matrix profile:
binding_sites <- rba_jaspar_sites(matrix_id = "MA0600.2")
```

------------------------------------------------------------------------

# TF flexible models (TFFMs) {#tf-flexible-models-tffms}

JASPAR also stores and assigns identifiers to TF flexible models (TFFMs). As with position frequency matrices (PFMs), you can search TFFMs or retrieve information and annotations using a TFFM identifier. TFFM IDs are versioned and follow the `base_id.version` format.

```{r tffm_search, eval=FALSE}
## Search TFFMs. This is a search function. Thus, what has been presented
## in the `Search Matrix Profiles` section also applies here:

## Get a list of all available TFFM profiles:
all_tffms <- rba_jaspar_tffm_search()

## Search FOX:
FOX_tffms <- rba_jaspar_tffm_search(term = "FOX")

## Transcription factors named FOXP3
FOXP3_tffms <- rba_jaspar_tffm_search(term = "FOXP3")

## Transcription factors in the insects taxonomic group
insects_tffms <- rba_jaspar_tffm_search(tax_group = "insects")
```

```{r tffm}
## Now that you have a TFFM ID, you can retrieve it
TFFM0056 <- rba_jaspar_tffm("TFFM0056.3")

str(TFFM0056)
```

------------------------------------------------------------------------

# How to Cite? {#citations}

To cite JASPAR (Please see <https://jaspar.elixir.no/faq/>):

-   Baydar Ovek D, et al. *JASPAR 2026: expansion of transcription factor binding profiles and integration of deep learning models*. Nucleic Acids Res. 2026;54(D1):D184-D193; <https://doi.org/10.1093/nar/gkaf1209>
- Khan, A. and Mathelier, A. *JASPAR RESTful API: accessing JASPAR data from any programming language*. Bioinformatics, 2017, <https://doi.org/10.1093/bioinformatics/btx804> 


To cite rbioapi:

-   Moosa Rezwani, Ali Akbar Pourfathollah, Farshid Noorbakhsh, rbioapi: user-friendly R interface to biologic web services' API, Bioinformatics, Volume 38, Issue 10, 15 May 2022, Pages 2952--2953, <https://doi.org/10.1093/bioinformatics/btac172>

------------------------------------------------------------------------

# Links {#links}

-   [This article in rbioapi documentation site](https://rbioapi.moosa-r.com/articles/rbioapi_jaspar.html "2.G: JASPAR & rbioapi")

-   [Functions references in rbioapi documentation site](https://rbioapi.moosa-r.com/reference/index.html#section-jaspar-rba-jaspar- "rbioapi reference")

-   [rbioapi vignette index](rbioapi.html "rbioapi: User-Friendly R Interface to Biologic Web Services' API")

------------------------------------------------------------------------

# Session info {#session-info}

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
