---
title: "2.E: Reactome & rbioapi"
author: "Moosa Rezwani"
description: >
 Connect to Reactome in R with rbioapi package. 
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    anchor_sections: true
    number_sections: true
    self_contained: true
    dev: png
    encoding: 'UTF-8'

vignette: >
  %\VignetteIndexEntry{2.E: Reactome & rbioapi}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r message=FALSE, include=FALSE, results="hide", setup, echo=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  tidy = FALSE,
  cache = FALSE,
  dev = "png",
  comment = "#>"
)
library(rbioapi)
rba_options(timeout = 30, skip_error = TRUE)
```


# Introduction {#introduction}

Directly quoting from [Reactome](https://reactome.org/):

> REACTOME is an open-source, open access, manually curated and peer-reviewed pathway database. Our goal is to provide intuitive bioinformatics tools for the visualization, interpretation and analysis of pathway knowledge to support basic and clinical research, genome analysis, modeling, systems biology and education. Founded in 2003, the Reactome project is led by Lincoln Stein of [OICR](https://oicr.on.ca/), Peter D'Eustachio of [NYULMC](https://nyulangone.org/), Henning Hermjakob of [EMBL-EBI](https://www.ebi.ac.uk/), and Guanming Wu of [OHSU](https://www.ohsu.edu/).
>
> (source: <https://reactome.org/what-is-reactome>)

Reactome provides two RESTful API services: Reactome content services and Reactome analysis services. In rbioapi, the naming schema is that any function which belongs to analysis services starts with rba_reactome_analysis\* . Other rba_reactome\_\* functions without the 'analysis' infix correspond to content services API.

Before continuing reading this article, it is a good idea to read [Reactome Data Model](https://reactome.org/documentation/data-model) page.

------------------------------------------------------------------------

# Reactome analysis services {#reactome-analysis-services}

This section mostly revolves around `rba_reactome_analysis()` function. So, naturally, we will start with that. As explained in the function's manual, you have considerable freedom in providing the main input for this function; You can supply an R object (as a data frame, matrix, or simple vector), a URL, or a local file path. Note that the type of analysis will be decided based on whether your input is 1-dimensional or 2-dimensional. This has been explained in detail in the manual of `rba_reactome_analysis()`, see that for more information.\
`rba_reactome_analysis()` is the API equivalent of Reactome's [analyse gene list](https://reactome.org/PathwayBrowser/#TOOL=AT) tool. You can see that the function's arguments correspond to what would you choose in the webpage's wizard.

```{r rba_reactome_analysis}
## 1 We create a simple vector with our genes
genes <- c(
  "p53", "BRCA1", "cdk2", "Q99835", "CDC42", "CDK1", "KIF23", "PLK1", "RAC2",
  "RACGAP1", "RHOA", "RHOB", "MSL1", "PHF21A", "INSR", "JADE2", "P2RX7",
  "CCDC101", "PPM1B", "ANAPC16", "CDH8", "HSPA1L", "CUL2", "ZNF302", "CUX1",
  "CYTH2", "SEC22C", "EIF4E3", "ROBO2", "CXXC1", "LINC01314", "ATP5F1"
)

## 2 We call reactome analysis with the default parameters
analyzed <- rba_reactome_analysis(
  input = genes,
  projection = TRUE,
  p_value = 0.01
)

## 3 As always, we use str() to inspect the resutls
str(analyzed, 1)

## 4 Note that in the summary element: (analyzed$summary)
### 4.a because we supplied a simple vector, the analysis type was: over-representation
### 4.b You need the token for other rba_reactome_analysis_* functions

## 5 Analsis results are in the pathways data frame:
```

```{r analysis_results, echo=FALSE}
if (utils::hasName(analyzed, "pathways") && is.data.frame(analyzed$pathways)) {
  DT::datatable(
    data = jsonlite::flatten(analyzed$pathways),
    options = list(
      scrollX = TRUE, 
      paging = TRUE,
      fixedHeader = TRUE,
      keys = TRUE,
      pageLength = 5
    )
  )
} else {
  print("Vignette building failed. It is probably because the web service was down during the building.")
}

```

As mentioned, some of `rba_reactome_analysis()`'s arguments correspond to the wizard of [analyse gene list](https://reactome.org/PathwayBrowser/#TOOL=AT) tool; Other arguments corresponds to the contents of "Filter your results" tab in the results page.

Having the analysis's token, you can retrieve the analysis results in many formats using `rba_reactome_analysis_pdf()` and `rba_reactome_analysis_download()`:

```{r rba_reactome_analysis_pdf/download, eval=FALSE}
# download a full pdf report
rba_reactome_analysis_pdf(
  token = analyzed$summary$token,
  species = 9606
)

# download the result in compressed json.gz format
rba_reactome_analysis_download(
  token = analyzed$summary$token,
  request = "results",
  save_to = "reactome_results.json"
)
```

Your token is only guaranteed to be stored for 7 days. After that, you can upload the JSON file you have downloaded using `rba_reactome_analysis_download` and get a token for that:

```{r rba_reactome_analysis_import, eval=FALSE}
re_uploaded <- rba_reactome_analysis_import(input = "reactome_results.json")
```

**Please Note:** Other services supported by rbioapi also provide Over-representation analysis tools. Please see the vignette article [Do with rbioapi: Over-Representation (Enrichment) Analysis in R](rbioapi_do_enrich.html) ([link to the documentation site](https://rbioapi.moosa-r.com/articles/rbioapi_do_enrich.html)) for an in-depth review.

## See also in Functions' manuals {#analysis_see-also-in-functions-manuals}

Some rbioapi Reactome analysis functions were not covered in this vignette, be sure to check their manuals:

-   `rba_reactome_analysis_mapping()`

-   `rba_reactome_analysis_species()`

-   `rba_reactome_analysis_token()`

------------------------------------------------------------------------

# Reactome contents services {#reactome-contents-services}

rbioapi functions that correspond to Reactome content services are those starting with rba_reactome\_\* but without "\_analysis" infix. These functions cover what you can do with objects in Reactome knowledge-base. In simpler terms, most -but not all of them- correspond to what you can find in [Reactome Pathway Browser](https://reactome.org/PathwayBrowser/ "Reactome Pathway Browser") and [search results](https://reactome.org/content/query?q=apoptosis&species=Homo+sapiens "Search results for apoptosis"). (e.g. a [pathway](https://reactome.org/content/detail/R-HSA-109581 "Apoptosis Homo sapiens"), a [reaction](https://reactome.org/content/detail/R-HSA-202939 "Caspase-mediated cleavage of E-Cadherin"), a [physical Entity](https://reactome.org/content/detail/R-HSA-350870 "Caspase-3 [cytosol]"), etc.)

## Search the Reactome knowledge-base {#search-the-reactome-knowledge-base}

If you do not already know the Reactome identifier of an entry, use `rba_reactome_search()` to search by name or descriptive text. The results can be limited by species, entry type, cellular compartment, and keyword. By default, matches are grouped by their result type. In the following example, we search for TP53 across four representative human entry types and request one match from each group.

```{r rba_reactome_search}
search_results <- rba_reactome_search(
  query = "TP53",
  species = "Homo sapiens",
  types = c("Protein", "Complex", "Reaction", "Pathway"),
  page_size = 1,
  force_filters = TRUE
)

str(search_results, 2)
```

The matching entries are stored in the `entries` column of the returned result groups. Below, we combine the four groups and display selected fields:

```{r rba_reactome_search_results, echo=FALSE}
if (
  is.list(search_results) &&
  utils::hasName(search_results, "results") &&
  is.data.frame(search_results$results) &&
  utils::hasName(search_results$results, "entries") &&
  length(search_results$results$entries) > 0
) {
  search_entries <- do.call(
    rbind,
    lapply(
      search_results$results$entries,
      function(entries) {
        entries[, c("id", "name", "type"), drop = FALSE]
      }
    )
  )
  search_entries$name <- gsub("<[^>]+>", "", search_entries$name)

  DT::datatable(
    data = search_entries,
    rownames = FALSE,
    options = list(
      dom = "t",
      paging = FALSE,
      ordering = FALSE
    )
  )
} else {
  print("Vignette building failed. It is probably because the web service was down during the building.")
}
```

## Retrieve any object from Reactome knowledge-base {#retrieve-any-object-from-reactome-knowledge-base}

Using `rba_reactome_query()`, you can retrieve any object to which Reactome has assigned a database or stable identifier. This includes proteins, reactions, pathways, species, people, and many other entries described in [Reactome's data schema](https://reactome.org/content/schema/DatabaseObject "Graph Database :: Data Schema"). A standard query accepts one or more identifiers. An enhanced query accepts one identifier and adds related regulations and catalysts; it also lets you control incoming relationships and disease-specific information, or request a shorter summary for a reference entity.

```{r rba_reactome_query_ex1}
## 1 Query a pathway entry
pathway <- rba_reactome_query(
  ids = "R-HSA-109581",
  enhanced = TRUE
)

## 2 As always we use str() to inspect the output's structure
str(pathway, 2)



## 3 Compare the result with the pathway's Reactome page
# https://reactome.org/content/detail/R-HSA-109581
```

```{r rba_reactome_query_ex2}
## 1 Query a reference entity and summarize its physical forms
protein <- rba_reactome_query(
  ids = 66247,
  enhanced = TRUE,
  summarize_reference_entity = TRUE
)

## 2 As always we use str() to inspect the output's structure
str(protein, 1)



## 3 Compare the result with the entry's Reactome page
# https://reactome.org/content/detail/66247

```

## Find Cross-Reference IDs in Reactome {#find-cross-reference-ids-in-reactome}

In the second example, we used Reactome's database identifier `66247` to query the CD40 reference entity. `rba_reactome_xref()` can map an external identifier, such as a gene symbol, to the corresponding Reactome reference entity.

```{r rba_reactome_xref}
## 1 Supply an HGNC symbol to find the corresponding Reactome database ID
xref_protein <- rba_reactome_xref("CD40")

## 2 As always, use str() to inspect the output's structure
str(xref_protein, 1)
```

Set `expanded = TRUE` to also retrieve other external identifiers associated with the reference entity and the stable identifiers of its physical forms. Expanded queries can accept a vector of identifiers; `page` and `page_size` select which supplied identifiers Reactome processes in each call. The optional database filter uses Reactome's database names.

```{r rba_reactome_xref_expanded}
## Retrieve the ENSEMBL cross-references and associated physical forms
xref_details <- rba_reactome_xref(
  "P36897",
  expanded = TRUE,
  db_filter = "ENSEMBL"
)

str(xref_details, 2)
```

## Map Cross-Reference IDs to Reactome {#map-cross-reference-ids-to-reactome}

While we are at the cross-reference topic, here is another useful resource. Using `rba_reactome_mapping` you can find the Reactome pathways or reactions which include your external ID:

```{r xref_mapping}
## 1 Again, consider CD40 protein:
xref_mapping <- rba_reactome_mapping(
  id = "CD40",
  resource = "hgnc",
  map_to = "pathways"
)
```

```{r xref_mapping_df, echo=FALSE}
if (is.data.frame(xref_mapping)) {
  DT::datatable(
    data = xref_mapping,
    options = list(
      scrollX = TRUE, 
      paging = TRUE,
      fixedHeader = TRUE,
      keys = TRUE,
      pageLength = 10
    )
  )
} else {
  print("Vignette building failed. It is probably because the web service was down during the building.")
}
```

------------------------------------------------------------------------

# See also in function manuals {#see-also-in-functions-manuals}

Several rbioapi Reactome content functions are not covered in this vignette. The following overview links them by purpose; see each function's manual for details and examples.

## Retrieve Reactome database information {#retrieve-reactome-database-information}

-   `rba_reactome_version()`: Return current Reactome version

-   `rba_reactome_diseases()`: Retrieve a list of disease annotated in Reactome.

-   `rba_reactome_species()`: Retrieve a list of species annotated in Reactome.

## General Mapping/Querying {#general-mapping-querying}

-   `rba_reactome_search()`: Search Reactome entries by text and optional filters.

-   `rba_reactome_query()`: Retrieve Reactome objects, optionally with enhanced relationships or a selected attribute.

-   `rba_reactome_mapping()`

-   `rba_reactome_xref()`: Map external identifiers to Reactome reference entities and, optionally, their other cross-references and physical forms.

## Things you can do with entities {#things-you-can-do-with-a-entities}

-   `rba_reactome_complex_list()`: Get a list of complexes that have your molecule in them.

-   `rba_reactome_complex_subunits()`: Get the list of subunits in your complex

-   `rba_reactome_participant_of()`: Get a list of Reactome sets and complexes that your entity (event, molecule, reaction, pathway etc.) is a participant in them.

-   `rba_reactome_entity_other_forms()`

## Things you can do with Events {#things-you-can-do-with-events}

-   `rba_reactome_event_ancestors()`

-   `rba_reactome_participants()`

-   `rba_reactome_pathways_events()`

-   `rba_reactome_event_ancestors()`

-   `rba_reactome_orthology()`

-   `rba_reactome_event_hierarchy()`: Retrieve a species's event or pathway hierarchy, optionally with analysis results added.

## Pathways {#pathways}

-   `rba_reactome_pathways_low()`

-   `rba_reactome_pathways_events()`

-   `rba_reactome_pathways_top()`

## Interactors {#interactors}

-   `rba_reactome_interactors_psicquic()`

-   `rba_reactome_interactors_static()`: Retrieve Reactome's static IntAct interaction details, summaries, or associated pathways.

## People {#peaple}

-   `rba_reactome_people_name()`

-   `rba_reactome_people_id()`

## Export diagrams and events {#export-diagrams-and-events}

-   `rba_reactome_exporter_diagram()`

-   `rba_reactome_exporter_overview()`

-   `rba_reactome_exporter_reaction()`

-   `rba_reactome_exporter_event()`

------------------------------------------------------------------------

# How to Cite? {#citations}

To cite Reactome, please reference one or more of the following publications
(see <https://reactome.org/cite>):

-   Ragueneau, E., Gong, C., Sinquin, P., Sevilla, C., Beavers, D., Grentner, A., ... D’Eustachio, P. (2026). The Reactome Knowledgebase 2026. *Nucleic Acids Research*, 54(D1), D673–D681. <https://doi.org/10.1093/nar/gkaf1223>.
-   Griss, J., Viteri, G., Sidiropoulos, K., Nguyen, V., Fabregat, A., & Hermjakob, H. (2020). ReactomeGSA—Efficient Multi-Omics Comparative Pathway Analysis. *Molecular & Cellular Proteomics*, 19(12), 2115–2125. <https://doi.org/10.1074/mcp.TIR120.002155>.
-   Fabregat, A., Korninger, F., Viteri, G., Sidiropoulos, K., Marin-Garcia, P., Ping, P., Wu, G., Stein, L., D'Eustachio, P., & Hermjakob, H. (2018). Reactome graph database: Efficient access to complex pathway data. *PLOS Computational Biology*, 14(1), e1005968. <https://doi.org/10.1371/journal.pcbi.1005968>.
-   Fabregat, A., Sidiropoulos, K., Viteri, G., Marin-Garcia, P., Ping, P., Stein, L., D'Eustachio, P., & Hermjakob, H. (2018). Reactome diagram viewer: Data structures and strategies to boost performance. *Bioinformatics*, 34(7), 1208–1214. <https://doi.org/10.1093/bioinformatics/btx752>.
-   Fabregat, A., Sidiropoulos, K., Viteri, G., Forner, O., Marin-Garcia, P., Arnau, V., D'Eustachio, P., Stein, L., & Hermjakob, H. (2017). Reactome pathway analysis: A high-performance in-memory approach. *BMC Bioinformatics*, 18(1), 142. <https://doi.org/10.1186/s12859-017-1559-2>.
-   Wu, G., & Haw, R. (2017). Functional Interaction Network Construction and Analysis for Disease Discovery. *Methods in Molecular Biology*, 1558, 235–253. <https://doi.org/10.1007/978-1-4939-6783-4_11>.

To cite rbioapi:

-   Moosa Rezwani, Ali Akbar Pourfathollah, Farshid Noorbakhsh, rbioapi: user-friendly R interface to biologic web services' API, Bioinformatics, Volume 38, Issue 10, 15 May 2022, Pages 2952--2953, <https://doi.org/10.1093/bioinformatics/btac172>

------------------------------------------------------------------------

# Links {#links}

-   [This article in rbioapi documentation site](https://rbioapi.moosa-r.com/articles/rbioapi_reactome.html "2.D: Reactome & rbioapi")

-   [Functions references in rbioapi documentation site](https://rbioapi.moosa-r.com/reference/index.html#section-reactome-analysis-services-rba-reactome- "rbioapi reference")

-   [rbioapi vignette index](rbioapi.html "rbioapi: User-Friendly R Interface to Biologic Web Services' API")

------------------------------------------------------------------------

# Session info {#session-info}

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
