---
title: "2.C: miEAA & rbioapi"
author: "Moosa Rezwani"
description: >
 Connect to miEAA in R with rbioapi package. 
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    anchor_sections: true
    number_sections: true
    self_contained: true
    dev: png
    encoding: 'UTF-8'

vignette: >
  %\VignetteIndexEntry{2.C: miEAA & rbioapi}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r message=FALSE, include=FALSE, results="hide", setup, echo=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  tidy = FALSE,
  cache = FALSE,
  dev = "png",
  comment = "#>"
)
library(rbioapi)
rba_options(timeout = 30, skip_error = TRUE)
```


# Introduction {#introduction}

The miRNA Enrichment Analysis and Annotation Tool ([miEAA](https://ccb-compute2.cs.uni-saarland.de/mieaa/ "miRNA Enrichment Analysis and Annotation Tool")) is a multi-species microRNA enrichment-analysis service provided by the [Chair for Clinical Bioinformatics at Saarland University](https://www.ccb.uni-saarland.de/). For more information, see the miEAA [website](https://ccb-compute2.cs.uni-saarland.de/mieaa/) or its [latest publication](https://doi.org/10.1093/nar/gkad392 "miEAA 2023: updates, new functional microRNA sets and improved enrichment visualizations").

------------------------------------------------------------------------

# First, find **enrichment categories** {.heading2}

Before performing enrichment analysis on a miRNA set, note that the supported **enrichment categories** depend on both the **miRNA type** (mature or precursor, without mixing the two) and the **species**. See the [miEAA integrated data sets](https://ccb-compute2.cs.uni-saarland.de/mieaa/downloads/) for details.

Thus, it is recommended to retrieve a list of possible enrichment categories that you may use:

```{r rba_mieaa_cats, eval=FALSE}
## A list of available enrichment categories for:

## mature human miRNA:
rba_mieaa_cats(mirna_type = "mature", species = 9606)

## precursor human miRNA
rba_mieaa_cats(mirna_type = "precursor", species = 9606)

## precursor zebrafish miRNA
rba_mieaa_cats(mirna_type = "precursor", species = "Danio rerio")
```

------------------------------------------------------------------------

# Submit an enrichment-analysis request to miEAA {#submit-enrichment-analysis-request-to-mieaa}

There are two approaches to do this, we will start with the simpler one.

## Approach 1: Using the wrapper function {#approach-1-using-the-wrapper-function}

Supply the arguments of `rba_mieaa_enrich()` as described in its manual. You must provide `test_set`, `mirna_type`, `test_type`, and `species`:

```{r rba_mieaa_enrich, message=TRUE}
## 1 We create a variable with our miRNAs' mature IDs
mirs <- c(
  "hsa-miR-20b-5p", "hsa-miR-144-5p", "hsa-miR-17-5p", "hsa-miR-20a-5p",
  "hsa-miR-222-3p", "hsa-miR-106a-5p", "hsa-miR-93-5p", "hsa-miR-126-3p",
  "hsa-miR-363-3p", "hsa-miR-302c-3p", "hsa-miR-374b-5p", "hsa-miR-18a-5p",
  "hsa-miR-548d-3p", "hsa-miR-135a-3p", "hsa-miR-558", "hsa-miR-130b-5p",
  "hsa-miR-148a-3p"
)

## 2a Perform enrichment analysis without limiting it to selected categories
mieaa_all <- rba_mieaa_enrich(
  test_set = mirs,
  mirna_type = "mature",
  test_type = "ORA",
  species = 9606
)

## 2b Limit the enrichment to selected data sets (enrichment categories)
mieaa_kegg <- rba_mieaa_enrich(
  test_set = mirs,
  mirna_type = "mature",
  test_type = "ORA",
  species = 9606,
  categories = "KEGG_mature"
)
```

```{r mieaa_kegg_table, echo=FALSE}
if (is.data.frame(mieaa_kegg)) {
  DT::datatable(
    data = mieaa_kegg,
    options = list(
      scrollX = TRUE, 
      paging = TRUE,
      fixedHeader = TRUE,
      keys = TRUE,
      pageLength = 10
    )
  )
} else {
  print("Vignette building failed. It is probably because the web service was down during the building.")
}
```

## Approach 2: Going step-by-step {#approach-2-going-step-by-step}

As stated before, `rba_mieaa_enrich()` is a wrapper function, meaning that it executes the following sequence of functions:

```{r rba_meaa_submit/status/results, eval=FALSE}
## 1 Submit enrichment request to miEAA
request <- rba_mieaa_enrich_submit(
  test_set = mirs,
  mirna_type = "mature",
  test_type = "ORA",
  species = 9606,
  categories = c("miRWalk_Diseases_mature", "miRWalk_Organs_mature")
)

## 2 check for job's running status
rba_mieaa_enrich_status(job_id = request$job_id)

## 3 If the job has completed, retrieve the results
results <- rba_mieaa_enrich_results(job_id = request$job_id)
```

**Please Note:** Other services supported by rbioapi also provide Over-representation analysis tools. Please see the vignette article [Do with rbioapi: Over-Representation (Enrichment) Analysis in R](rbioapi_do_enrich.html) ([link to the documentation site](https://rbioapi.moosa-r.com/articles/rbioapi_do_enrich.html)) for an in-depth review.

------------------------------------------------------------------------

# Convert miRNA identifiers {#convert-mirna-accessions}

miEAA uses miRBase version 22 identifiers. Use `rba_mieaa_convert_version()` to convert miRNA identifiers between supported miRBase versions. miEAA also distinguishes mature and precursor miRNA identifiers; use `rba_mieaa_convert_type()` to convert between these identifier types.

------------------------------------------------------------------------

# How to Cite? {#citations}

To cite miEAA (see <https://ccb-compute2.cs.uni-saarland.de/mieaa/>):

-   Ernesto Aparicio-Puerta, Pascal Hirsch, Georges P. Schmartz, Fabian Kern, Tobias Fehlmann, Andreas Keller, miEAA 2023: updates, new functional microRNA sets and improved enrichment visualizations, Nucleic Acids Research, Volume 51, Issue W1, 5 July 2023, Pages W319--W325, <https://doi.org/10.1093/nar/gkad392>

To cite rbioapi:

-   Moosa Rezwani, Ali Akbar Pourfathollah, Farshid Noorbakhsh, rbioapi: user-friendly R interface to biologic web services' API, Bioinformatics, Volume 38, Issue 10, 15 May 2022, Pages 2952--2953, <https://doi.org/10.1093/bioinformatics/btac172>

------------------------------------------------------------------------

# Links {#links}

-   [This article on the rbioapi documentation site](https://rbioapi.moosa-r.com/articles/rbioapi_mieaa.html "2.C: miEAA & rbioapi")

-   [Function references on the rbioapi documentation site](https://rbioapi.moosa-r.com/reference/index.html#section-mieaa-rba-mieaa- "rbioapi reference")

-   [rbioapi vignette index](rbioapi.html "rbioapi: User-Friendly R Interface to Biologic Web Services' API")

------------------------------------------------------------------------

# Session info {#session-info}

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
