---
title: "2.A: Enrichr & rbioapi"
author: "Moosa Rezwani"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    anchor_sections: true
    number_sections: true
    self_contained: true
    dev: png
    encoding: 'UTF-8'

vignette: >
  %\VignetteIndexEntry{2.A: Enrichr & rbioapi}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r message=FALSE, include=FALSE, results="hide", setup, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      eval = TRUE,
                      message = FALSE,
                      warning = FALSE,
                      collapse = TRUE,
                      tidy = FALSE,
                      cache = FALSE,
                      dev = "png",
                      comment = "#>")
library(rbioapi)
rba_options(timeout = 600, skip_error = TRUE)
```

------------------------------------------------------------------------

# Introduction {#introduction .heading2}

[Enrichr](https://maayanlab.cloud/Enrichr) is a gene-set enrichment analysis tool developed in the [Ma'ayan lab](https://icahn.mssm.edu/research/labs/maayan-laboratory).

# Gene set library concept in Enrichr {#gene-set-library-concept-in-enrichr .heading2}

Directly quoting from Enrichr's help page;

> A *gene set library* is a set of related gene sets or enrichment terms. Each enrichment term in Enrichr's results pages is organized by its gene set library. These libraries have been constructed from many sources such as published studies and major biological and biomedical online databases. Others have been created for and only available through Enrichr. For example, the ChEA 2015 library is a set of functional terms representing transcription factors profiled by ChIP-seq in mammalian cells. Each term is associated with a collection of putative targets inferred from the peaks identified in each ChIP-seq study.
>
> (source: <https://maayanlab.cloud/Enrichr/help#background>)

To get a list of available Enrichr libraries, use:

```{r enrichr_libs}
enrichr_libs <- rba_enrichr_libs()
```

In the returned data frame, you can find the names of available Enrichr libraries in "libraryName" column:

```{r enrichr_libs_df, echo=FALSE}
if (is.data.frame(enrichr_libs)) {
  DT::datatable(data = enrichr_libs,
              options = list(scrollX = TRUE, 
                             paging = TRUE,
                             fixedHeader = TRUE,
                             keys = TRUE,
                             pageLength = 10))
} else {
  print("Vignette building failed. It is probably because the web service was down during the building.")
}
```

------------------------------------------------------------------------

# Enrichment analysis using Enrichr {#enrichment-analysis-using-enrichr .heading2}

To enrich your gene-set with Enrichr using rbioapi, you can take two approaches. we will begin with the simple one.

## Approach 1: Using the Wrapper function {#approach-1-using-the-wrapper-function .heading3}

Just fill the arguments of `rba_enrichr` according to the function's manual; Simply supply your gene-set as a character vector and select the libraries.

```{r approach_1}
# 1 We create a variable with our genes' NCBI IDs
genes <- c("p53", "BRCA1", "cdk2", "Q99835", "CDC42","CDK1","KIF23","PLK1",
           "RAC2","RACGAP1","RHOA","RHOB", "PHF14", "RBM3", "MSL1")

# 2.a Enrich your genes using "MSigDB_Hallmark_2020" library
enrichr_msig_hallmark <- rba_enrichr(gene_list = genes,
                                     gene_set_library = "MSigDB_Hallmark_2020")
# 2.b Maybe you want to enrich using every library that contains the word "msig":
enrichr_msig <- rba_enrichr(gene_list = genes,
                            gene_set_library = "msig",
                            regex_library_name = TRUE)
# 2.c Or maybe you want to enrich using every library available at Enrichr:
# enrichr_all <- rba_enrichr(gene_list = genes,
#                            gene_set_library = "all")
```

Note that when only one Enrichr library is selected, a data frame with enrichment analysis result will be returned:

```{r approach_1_single}
str(enrichr_msig_hallmark)
```

But when multiple libraries have been selected, the function's output will be a list where each element is a data frame corresponding to one of the selected libraries:

```{r approach_1_multi}
str(enrichr_msig, 1)
```

## Approach 2: Going step-by-step {#approach-2-going-step-by-step .heading3}

As you can see in `rba_enrichr()`'s name, it is a wrapper function. It basically executes the following sequence of functions:

```{r approach_2}
# 1 Get a list of available Enrichr libraries
libs <- rba_enrichr_libs(store_in_options = TRUE)

# 2 Submit your gene-set to enrichr
list_id <- rba_enrichr_add_list(gene_list = genes)

# 3 Perform Enrichment analysis with your uploaded gene-set
enriched <- rba_enrichr_enrich(user_list_id = list_id$userListId,
                               gene_set_library = "Table_Mining_of_CRISPR_Studies")

## As always, use str() to see what you have:
str(enriched, 1)
```

# See also in Functions' manuals {#see-also-in-functions-manuals .heading2}

Some rbioapi Enrichr functions were not covered in this vignette, be sure to check their manuals:

-   `rba_enrichr_gene_map()`

-   `rba_enrichr_view_list()`

------------------------------------------------------------------------

# How to Cite? {#Citations}

To cite Enrichr (Please see <https://maayanlab.cloud/Enrichr/help#terms>):

-   Chen, E.Y., Tan, C.M., Kou, Y. *et al.* Enrichr: interactive and collaborative HTML5 gene list enrichment analysis tool. *Bioinformatics* **14,** 128 (2013). <https://doi.org/10.1186/1471-2105-14-128>

-   Maxim V. Kuleshov, Matthew R. Jones, Andrew D. Rouillard, Nicolas F. Fernandez, Qiaonan Duan, Zichen Wang, Simon Koplev, Sherry L. Jenkins, Kathleen M. Jagodnik, Alexander Lachmann, Michael G. McDermott, Caroline D. Monteiro, Gregory W. Gundersen, Avi Ma'ayan, Enrichr: a comprehensive gene set enrichment analysis web server 2016 update, *Nucleic Acids Research*, Volume 44, Issue W1, 8 July 2016, Pages W90--W97, <https://doi.org/10.1093/nar/gkw377>

-   Xie, Z., Bailey, A., Kuleshov, M. V., Clarke, D. J. B., Evangelista, J. E., Jenkins, S. L., Lachmann, A., Wojciechowicz, M. L., Kropiwnicki, E., Jagodnik, K. M., Jeon, M., & Ma'ayan, A. (2021). Gene set knowledge discovery with Enrichr. *Current Protocols*, 1, e90. doi: <https://doi.org/10.1002/cpz1.90>

To cite rbioapi:

-   Moosa Rezwani, Ali Akbar Pourfathollah, Farshid Noorbakhsh, rbioapi: User-Friendly R Interface to Biologic Web Services' API, *Bioinformatics*, 2022;, btac172, <https://doi.org/10.1093/bioinformatics/btac172>

------------------------------------------------------------------------

# Links {.heading2}

-   [This article in rbioapi documentation site](https://rbioapi.moosa-r.com/articles/rbioapi_enrichr.html "https://rbioapi.moosa-r.com/articles/rbioapi_enrichr.html")

-   [Functions references in rbioapi documentation site](https://rbioapi.moosa-r.com/reference/index.html#section-enrichr-rba-enrichr- "rbioapi reference")

-   [rbioapi vignette index](rbioapi.html "rbioapi: User-Friendly R Interface to Biologic Web Services' API")

------------------------------------------------------------------------

# Session info {#session-info .heading2}

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
