---
title: "2.F: STRING & rbioapi"
author: "Moosa Rezwani"
description: >
 Connect to STRING in R with rbioapi package.
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    anchor_sections: true
    number_sections: true
    self_contained: true
    dev: png
    encoding: 'UTF-8'

vignette: >
  %\VignetteIndexEntry{2.F: STRING & rbioapi}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r message=FALSE, include=FALSE, results="hide", setup, echo=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  tidy = FALSE,
  cache = FALSE,
  dev = "png",
  comment = "#>"
)
library(rbioapi)
rba_options(timeout = 30, skip_error = TRUE)
```


# Introduction {#introduction}

STRING is a comprehensive database of known and predicted protein-protein
associations. It integrates direct physical interactions and indirect functional
associations from experimental evidence, curated databases, text mining,
co-expression, and computational predictions. rbioapi provides a consistent R
interface to STRING's mapping, network, homology, annotation, and enrichment
API resources.

------------------------------------------------------------------------

# The `species` argument {#note-about-species-argument}

Most rbioapi STRING functions accept a `species` argument containing an
[NCBI Taxonomy identifier](https://www.ncbi.nlm.nih.gov/taxonomy/), such as
9606 for human. Supplying it is recommended because it helps STRING resolve
ambiguous identifiers.

For the network, interaction-partner, homology, PPI-enrichment, network-image,
and enrichment-image functions, `species` is required when the input contains
more than 10 unique IDs. Mapping, functional-enrichment, and
functional-annotation requests do not have this local restriction. The limit is
based on unique IDs rather than the length of the original vector.

------------------------------------------------------------------------

# Map your IDs to STRING IDs {#map-your-ids-to-string-ids}

Although STRING recognizes a variety of identifiers, mapping them to STRING IDs
before using other rbioapi STRING functions makes subsequent requests faster
and less ambiguous.

```{r rba_string_map_ids}
## 1. Create a vector of protein identifiers
proteins <- c(
  "p53", "BRCA1", "cdk2", "Q99835", "CDC42", "CDK1", "KIF23",
  "PLK1", "RAC2", "RACGAP1", "RHOA", "RHOB", "PHF14", "RBM3"
)

## 2. Map the protein identifiers
proteins_mapped_df <- rba_string_map_ids(ids = proteins, species = 9606)
```

```{r rba_string_map_ids_results, echo=FALSE}
if (is.data.frame(proteins_mapped_df)) {
  proteins_mapped_df$annotation <- 
    paste0(
      substr(
        x = proteins_mapped_df$annotation,
        start = 1,
        stop = 50
      ),
      "..."
    )
  
  DT::datatable(
    data = proteins_mapped_df,
    options = list(
      scrollX = TRUE, 
      paging = TRUE,
      fixedHeader = TRUE,
      keys = TRUE,
      pageLength = 5
    )
  )
} else {
  print("Vignette building failed. It is probably because the web service was down during the building.")
}
```

```{r rba_string_map_ids2, eval=is.data.frame(proteins_mapped_df) && hasName(proteins_mapped_df, "stringId")}

## 3. Use the mapped STRING IDs in the remaining examples
proteins_mapped <- proteins_mapped_df$stringId
```

```{r rba_string_map_ids3, eval=!exists("proteins_mapped"), echo=!exists("proteins_mapped")}

# Fallback for vignette building
# if previous step failed due to communication error
  proteins_mapped <- c(
  "9606.ENSP00000269305", "9606.ENSP00000418960", "9606.ENSP00000266970",
  "9606.ENSP00000249373", "9606.ENSP00000497733", "9606.ENSP00000378699", 
  "9606.ENSP00000260363", "9606.ENSP00000300093", "9606.ENSP00000249071",
  "9606.ENSP00000404190", "9606.ENSP00000400175", "9606.ENSP00000272233", 
  "9606.ENSP00000385795", "9606.ENSP00000365950"
)
```
------------------------------------------------------------------------

# Get interaction network of a protein set {#get-interaction-network-of-a-protein-set}

You can retrieve interactions among the proteins in your set, together with the
scores for each STRING evidence channel. Use `required_score` to set the minimum
interaction score and `network_type` to select functional or physical
associations.

See the Value section of the `rba_string_interactions_network()` manual for
descriptions of the returned columns.

```{r rba_string_interactions_network}
int_net <- rba_string_interactions_network(
  ids = proteins_mapped,
  species = 9606,
  required_score = 500
)
```

```{r rba_string_interactions_network_results, echo=FALSE}
if (is.data.frame(int_net)) {
  DT::datatable(
    data = int_net,
    options = list(
      scrollX = TRUE, 
      paging = TRUE,
      fixedHeader = TRUE,
      keys = TRUE,
      pageLength = 5
    )
  )
} else {
  print("Vignette building failed. It is probably because the web service was down during the building.")
}
```

------------------------------------------------------------------------

# Get interaction partners of a protein set {#get-interaction-partners-of-a-protein-set}

The previous example returned interactions among the input proteins. To retrieve
proteins that interact with one or more input proteins, including proteins
outside the input set, use `rba_string_interaction_partners()`:

```{r rba_string_interaction_partners}
## This example uses one protein (CD40), but `ids` can contain multiple proteins
int_partners <- rba_string_interaction_partners(
  ids = "9606.ENSP00000361359",
  species = 9606,
  required_score = 900
)
```

```{r rba_string_interaction_partners_results, echo=FALSE}
if (is.data.frame(int_partners)) {
  DT::datatable(
    data = int_partners,
    options = list(
      scrollX = TRUE, 
      paging = TRUE,
      fixedHeader = TRUE,
      keys = TRUE,
      pageLength = 5
    )
  )
} else {
  print("Vignette building failed. It is probably because the web service was down during the building.")
}
```

------------------------------------------------------------------------

# Get network image of a protein set {#get-network-image-of-a-protein-set}

STRING can render the interaction network as a static image. The
`rba_string_network_image()` function supports several options for controlling
the network contents and appearance; see the function manual for details.

```{r rba_string_network_image_ex1, fig.show='hide'}
## Example 1:
graph_ppi1 <- rba_string_network_image(
  ids = proteins_mapped,
  image_format = "image",
  species = 9606,
  save_image = FALSE,
  required_score = 500,
  network_flavor = "confidence"
)
```

```{r rba_string_network_image_ex1_image, echo=FALSE, fig.cap="Network images - Example 1", fig.align='center', fig.width=7}
if (is.array(graph_ppi1)) {
  grid::grid.raster(graph_ppi1, just = "center")
} else {
  print("Vignette building failed. It is probably because the web service was down during the building.")
}
```

```{r rba_string_network_image_ex2, fig.show='hide'}
## Example 2:
graph_ppi2 <- rba_string_network_image(
  ids = proteins_mapped,
  image_format = "image",
  species = 9606,
  save_image = FALSE,
  required_score = 500,
  add_color_nodes = 5,
  add_white_nodes = 5,
  network_flavor = "actions"
)
```

```{r rba_string_network_image_ex2_image, echo=FALSE, fig.cap="Network images - Example 2", fig.align='center', fig.width=7}
if (is.array(graph_ppi2)) {
  grid::grid.raster(graph_ppi2, just = "center")
} else {
  print("Vignette building failed. It is probably because the web service was down during the building.")
}
```

------------------------------------------------------------------------

# Enrichment analysis using STRING {#enrichment-using-string}

STRING supports two complementary enrichment analyses: functional enrichment
and protein-protein interaction enrichment. See the
[STRING publication](https://doi.org/10.1093/nar/gkac1000) for methodological
details.

## Functional enrichment {#functional-enrichment}

Functional enrichment tests the supplied protein set against annotation
resources including Gene Ontology, KEGG pathways, UniProt Keywords, PubMed
publications, Pfam domains, InterPro domains, and SMART domains. The results
include raw p-values and false discovery rates. See the
[current STRING API documentation](https://string-db.org/help/api/#getting-functional-enrichment)
for details.

```{r rba_string_enrichment}
enriched <- rba_string_enrichment(
  ids = proteins_mapped,
  species = 9606
)
```

By default, results are split by category into a list of data frames. We can
inspect that structure with `str()`:

```{r rba_string_enrichment_str}
str(enriched, max.level = 1)
```

Let us see the "KEGG" results as an example. Below, we can see which terms of the [KEGG pathways database](https://www.genome.jp/kegg/pathway.html) were over-represented:

```{r rba_string_enrichment_results, echo=FALSE}
if (utils::hasName(enriched, "KEGG") && is.data.frame(enriched$KEGG)) {
  DT::datatable(
    data = enriched$KEGG,
    options = list(
      scrollX = TRUE, 
      paging = TRUE,
      fixedHeader = TRUE,
      keys = TRUE,
      pageLength = 5
    )
  )
} else {
  print("Vignette building failed. It is probably because the web service was down during the building.")
}
```

**Please Note:** Other services supported by rbioapi also provide Over-representation analysis tools. Please see the vignette article [Do with rbioapi: Over-Representation (Enrichment) Analysis in R](rbioapi_do_enrich.html) ([link to the documentation site](https://rbioapi.moosa-r.com/articles/rbioapi_do_enrich.html)) for an in-depth review.

## Functional enrichment plot {#functional-enrichment-plot}

The enrichment-figure endpoint summarizes one enrichment category as a plot.
Use `rba_string_enrichment_image()` to control the category, grouping, palette,
ranking variable, output format, and number of displayed terms. This example
visualizes the KEGG enrichment for the same protein set:


```{r rba_string_enrichment_plot, fig.show='hide'}
graph_enrich <- rba_string_enrichment_image(
  ids = proteins_mapped,
  species = 9606,
  category = "KEGG",
  image_format = "image",
  save_image = FALSE,
  group_by_similarity = 0.6
)
```

```{r rba_string_enrichment_plot_image, echo=FALSE, fig.cap="Visualization of enrichment analysis results", fig.align='center', fig.width=7}
if (is.array(graph_enrich)) {
  grid::grid.raster(graph_enrich, just = "center")
} else {
  print("Vignette building failed. It is probably because the web service was down during the building.")
}
```

## Protein-protein interaction enrichment {#protein-protein-interaction-enrichment}

PPI enrichment tests whether the input network contains more interactions than
expected for a similarly sized protein set drawn from the background proteome.
It can therefore identify evidence of functional relatedness without relying on
annotation enrichment.

```{r rba_string_enrichment_ppi}
rba_string_enrichment_ppi(
  ids = proteins_mapped,
  species = 9606
)
```

------------------------------------------------------------------------

# Get functional annotations {#get-functional-annotations}

STRING maps proteins to multiple annotation resources. Use
`rba_string_annotations()` to retrieve all annotations assigned to the input
proteins rather than only the subset returned by enrichment analysis. PubMed
annotations are excluded by default because they can make the response very
large.

```{r rba_string_annotations, eval=FALSE}
annotations <- rba_string_annotations(
  ids = "9606.ENSP00000269305",
  species = 9606
)

## Annotation responses can be large, so the result is not shown here
```

------------------------------------------------------------------------

# See also in Functions' manuals {#see-also-in-functions-manuals}

Some rbioapi STRING functions were not covered in this vignette, please check their manuals:

-   `rba_string_homology_intra()`

-   `rba_string_homology_inter()`

-   `rba_string_version()`

------------------------------------------------------------------------

# How to cite {#citations}

To cite STRING (Please see <https://string-db.org/cgi/about?footer_active_subpage=references>):

-   Damian Szklarczyk, Rebecca Kirsch, Mikaela Koutrouli, Katerina Nastou, Farrokh Mehryary, Radja Hachilif, Annika L Gable, Tao Fang, Nadezhda T Doncheva, Sampo Pyysalo, Peer Bork, Lars J Jensen, Christian von Mering, The STRING database in 2023: protein–protein association networks and functional enrichment analyses for any sequenced genome of interest, Nucleic Acids Research, Volume 51, Issue D1, 6 January 2023, Pages D638–D646, <https://doi.org/10.1093/nar/gkac1000>

To cite rbioapi:

-   Moosa Rezwani, Ali Akbar Pourfathollah, Farshid Noorbakhsh, rbioapi: user-friendly R interface to biologic web services' API, Bioinformatics, Volume 38, Issue 10, 15 May 2022, Pages 2952--2953, <https://doi.org/10.1093/bioinformatics/btac172>

------------------------------------------------------------------------

# Links {#links}

-   [This article in rbioapi documentation site](https://rbioapi.moosa-r.com/articles/rbioapi_string.html "2.E: STRING & rbioapi")

-   [Functions references in rbioapi documentation site](https://rbioapi.moosa-r.com/reference/index.html#section-string-rba-string- "rbioapi reference")

-   [rbioapi vignette index](rbioapi.html "rbioapi: User-Friendly R Interface to Biologic Web Services' API")

------------------------------------------------------------------------

# Session info {#session-info}

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
